Making Sequential Programs Deterministic

### 2019-04-01 design.tex
formatting
Edits in section 2.3

\begin{wrapfigure}{R}{0.3\textwidth}
    \begin{center}
        \includegraphics[scale=0.37]{detHierarchy}
    \end{center}
    \caption{Determinism type qualifier hierarchy}
    \label{fig:determinism-hierarchy}
%\todo{Drop the at-signs (@).}
%\todo{The text is weirdly small for the box.
%  Probably redraw this.}
\end{wrapfigure}

@@

A collection of type \<OrderNonDet> has special properties, including the following

\begin{enumerate}
\item
The individual elements retrieved from it have type \<NonDet>.  This
affects access, iteration, searching, and more.
(Rules \rulename{iteration} and \rulename{next element} in Figure~\ref{fig-ordernondet-rules})
\item
Size-related operations return a deterministic result.  This also affects
queries of whether an iterator has more elements.
\item
If the collection is sorted, or its elements are placed in a collection
that does sorting, the result is deterministic.
\end{enumerate}

To state the first point more formally, the 
typical type for list access, such as \<get>, is \\
$$\forall \tau . \<List>\angles{\tau} \times \<int> \rightarrow \tau$$
but this type is incorrect for \ourTypeSystem.  Each of the following is a 
$$
\forall \kappa\ \beta . \<List>\angles{\tau} \times \<int> \rightarrow \tau \\
\forall \tau . \<List>\angles{\tau} \times \<int> \rightarrow \tau \\
\forall \tau . \<List>\angles{\tau} \times \<int> \rightarrow \tau
$$

but we need to change that.  If either argument is \<OrderNonDet> or \<NonDet>, then the
result is \<NonDet>.
If both arguments are \<Det>, the result is $\tau$ as usual.
\todo{In that case, we know that $\tau = \<Det> \beta$.}

@@

\OurTypeSystem supports three types of polymorphism:  type
polymorphism, basetype polymorphism, and qualifier polymorphism.
These apply to both classes and methods.
\todo{Greekify the following}
\begin{itemize}
\item
In \theDeterminismCheckerImplementation,
type polymorphism is handled by Java's generics mechanism, which
\theDeterminismChecker fully supports, including class and method generics,
inference, etc.
Given the Java declaration
\begin{Verbatim}
<T> T identity(T arg) { return arg; }
\end{Verbatim}
the type of \<identity(arg)> is the same as the type of
\<arg>, and a programmer can write a type such as \codeid{List<@NonDet
  Integer>} (but see below for details about collections).
\item
Basetype polymorphism is enabled by writing a type qualifier on a use of a
type variable, which overrides the type qualifier at the instantiation
site.
For example, a \<choose> operation on sets could be defined in Java as
\begin{Verbatim}
class ChooseableSet<T> implements Set<T> {
  // return an arbitrary element from the set
  @NonDet T choose();
}
\end{Verbatim}
\item
Java does not provide a syntax that can be used for qualifier polymorphism,
so we use a special type qualifier name, \<@PolyDet>.
A qualifier-polymorphic method \<m> with signature $\forall \kappa . \kappa\ \<int> \times \<Det boolean> \rightarrow
\kappa\ \<String>$ would be written in our dialect of Java as
\begin{Verbatim}
@PolyDet String m(@PolyDet int, @Det boolean)
\end{Verbatim}
Each use of \<@PolyDet> stands for a use of the qualifier variable
$\kappa$, and there is no need to declare the qualifier variable $\kappa$.
\TheDeterminismChecker currently supports
qualifier
polymorphism on methods but not on classes.
Therefore, \<@PolyDet> may be written in methods (signatures and bodies)
but not on fields.
\end{itemize}

>>>>

%% Reinstate wrapfigure later.  But, if it spans pages, every following
%% page to the end of the document is narrowed to accommodate it.
% \begin{wrapfigure}{R}{0.3\textwidth}
\begin{figure}
    \begin{center}
        \includegraphics[scale=0.37]{detHierarchy}
    \end{center}
    \caption{Determinism type qualifier hierarchy}
    \label{fig:determinism-hierarchy}
\end{figure}
% \end{wrapfigure}

@@

A collection of type \<OrderNonDet> has special properties, including the following.

\begin{enumerate}
\item
The individual elements retrieved from it have type \<NonDet>.  This
affects access, iteration, searching, etc.
(Rules \rulename{iteration} and \rulename{next element} in Figure~\ref{fig-ordernondet-rules})
\item
Size-related operations return a deterministic result.  This also affects
queries of whether an iterator has more elements.
\item
If the collection is sorted, or its elements are placed in a collection
that does sorting, the result is deterministic.
\end{enumerate}

To state the first point more formally, the 
typical type for a list access operation, such as \<get>, is
$$\forall \tau . \<List>\angles{\tau} \times \<Det Int> \rightarrow \tau$$
but this type is incorrect for \ourTypeSystem.  Each of the following is a
correct (but overly restrictive) type for \<get>:
\begin{eqnarray*}
\forall \kappa\ \beta . \<NonDet>\ \<List>\angles{\kappa\ \beta} \times \<Det Int>
& \strut\hspace{-1em} \rightarrow \<NonDet>\ \beta
& \mbox{\quad in this case $\kappa = \<NonDet>$} \\
\forall \kappa\ \beta . \<OrderNonDet>\ \<List>\angles{\kappa\ \beta} \times \<Det Int>
& \strut\hspace{-1em} \rightarrow \<NonDet>\ \beta
& \mbox{\quad in this case $\kappa \in \{ \<OrderNonDet>, \<Det> \} $} \\
\forall \kappa\ \beta . \<Det>\ \<List>\angles{\kappa\ \beta} \times \<Det Int>
& \strut\hspace{-1em} \rightarrow \<Det>\ \beta\phantom{\<Non>}
& \mbox{\quad in this case $\kappa = \<Det>$}
\end{eqnarray*}

\noindent
and no type introduced so far can express this polymorphism.
(The typing is actually even more complex, because if \emph{either}
argument to \<get> is \<OrderNonDet> or \<NonDet>, then the
result is \<NonDet>.  The above examples only show the case where the index
is deterministic.)
The typical type fro the list access operation only applies if both
arguments are \<Det>,
and our typing must have that as a special case, which is the first line above.

@@

\todo{Cut parts of the below; show the rest in Greek.}

\OurTypeSystem supports both basetype polymorphism and qualifier polymorphism.
When both are applied, it gives the effect of type polymorphism.
Polymorphism applies to both classes and methods.

\todo{Greekify the following}
\begin{itemize}
\item
Type polymorphism has the usual semantics.  For example, the type of the
identity function is $\forall \tau . \tau \rightarrow \tau$, which can be
equivalently written as
$\forall \kappa \ \beta . \kappa \ \beta \rightarrow \kappa \ \beta$.
\item
Basetype polymorphism lets the basetype vary without applies a

An example of a type-polymorphic method is identity

In \theDeterminismCheckerImplementation,
type polymorphism is handled by Java's generics mechanism, which
\theDeterminismChecker fully supports, including class and method generics,
inference, etc.
Given the Java declaration
\begin{Verbatim}
<T> T identity(T arg) { return arg; }
\end{Verbatim}
the type of \<identity(arg)> is the same as the type of
\<arg>, and a programmer can write a type such as \codeid{List<@NonDet
  Integer>} (but see below for details about collections).
\item
Basetype polymorphism is enabled by writing a type qualifier on a use of a
type variable, which overrides the type qualifier at the instantiation
site.
For example, a \<choose> operation on sets could be defined in Java as
\begin{Verbatim}
class ChooseableSet<T> implements Set<T> {
  // return an arbitrary element from the set
  @NonDet T choose();
}
\end{Verbatim}
\item
Java does not provide a syntax that can be used for qualifier polymorphism,
so we use a special type qualifier name, \<@PolyDet>.
A qualifier-polymorphic method \<m> with signature $\forall \kappa . \kappa\ \<int> \times \<Det boolean> \rightarrow
\kappa\ \<String>$ would be written in our dialect of Java as
\begin{Verbatim}
@PolyDet String m(@PolyDet int, @Det boolean)
\end{Verbatim}
Each use of \<@PolyDet> stands for a use of the qualifier variable
$\kappa$, and there is no need to declare the qualifier variable $\kappa$.
\TheDeterminismChecker currently supports
qualifier
polymorphism on methods but not on classes.
Therefore, \<@PolyDet> may be written in methods (signatures and bodies)
but not on fields.
\end{itemize}
